**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**

### 2016-10-24 submission/Westcott_OptiClust_mSystems_2016.Rmd
Edit Methods section

***Sequence data and processing steps.***
To evaluate the OptiClust and the other algorithms we created two synthetic sequence collections and four sequence collections generated from previously published studies. The V4 region of the 16S rRNA gene was used from all datasets because it is a popular region that can be fully sequenced with two-fold coverage using the commonly used MiSeq sequencer from Illumina [@Kozich2013]. The method for generating the simulated datasets followed the approach used by Kopylova et al. [-@Kopylova2016] and Schloss [@Schloss2016]. Briefly, we randomly selected 10,000 uniques V4 fragments from 16S rRNA gene sequences that were unique from the SILVA non-redundant database [@Pruesse2007]. A community with an even relative abundance profile was generated by specifying that each sequence had a frequency of 100 reads. A community with a staggered relative abundance profile was generated by specifying that the abundance of each sequence was a randomly drawn integer sampled from a uniform distribution between 1 and 200. Sequence collections collected from human feces [@Baxter2016], murine feces [@Schloss2012], soil [@Johnston2016], and seawater [@Henson2016] were used to characterize the algorithms' performance with natural communities. These sequence collections were all generated using paired 150 or 250 nt reads of the V4 region. We re-processed all of the reads using a common analysis pipeline that included quality score-based error correction [@Kozich2013], alignment against a SILVA reference database [@Pruesse2007; @Schloss2010], screening for chimeras using UCHIME [@Edgar2011], and classification using a naive Bayesian classifier with the RDP training set [@Wang2007].

***Implementation of clustering algorithms.***
In addition to the OptiClust algorithm we evaluated ten different *de novo* clustering algorithms. These included three hierarchical algorithms, average neighbor (AN), nearest neighbor (NN), and furthest neighbor (FN), which are implemented in mothur [v.1.39.0; @Schloss2009]. Seven heuristic methods were also used including abundance-based greedy clustering (UAGC) and distance-based greedy clustering (UDGC) as implemented in USEARCH [v.6.1; @Edgar2010], abundance-based greedy clustering (VAGC) and distance-based greedy clustering (VDGC) as implemented in VSEARCH [v.X.X.X; @Rognes2015], OTUClust [v.X.X.X; @XXXXXX], SumaClust [v.X.X.X; @XXXXXX], and Swarm [v.2.1.1; @Mah2014]. With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments. To judge the quality of the Swarm-generated OTU assignments we calculated the MCC value using thresholds incremented by 1% between 0 and 5% and selected the threshold that provided the optimal MCC value [@Westcott2015]. We also assessed the ability of a previously describe heuristic to cluster sequences using the UDGC and OptiClust algorithms. In this heuristic sequences are split into bacterial families based on sequence classification and clustered within the taxonomic family [@Westcott20XX]. Finally, for our benchmarking analysis, we evaluate the memory and time requirements when using 8 processors with the UDGC and VDGC algorithms and with the taxonomic splitting heuristic.

***Benchmarking.*** We evaluated the quality of the sequence clustering, reproducibility of the clustering, the speed of clustering, and the amount of memory required to complete the clustering. To assess the quality of the clusters generated by each method, we counted the cells within a confusion matrix that indicated how well the clusterings represented the distances between the pair of sequences [@Schloss2011Assessing]. Pairs of sequences that were in the same OTU and had a distance less than 3% were true positives (TPs), those that were in different OTUs and had a distance greater than 3% were true negatives (TNs), those that were in the same OTU and had a distance greater than 3% were false positives (FPs), and those that were in different OTUs and had a distance less than 3% were false negatives (FNs). To synthesize the matrix into a single metric we used the Matthew's Correlation Coefficient, F1 score, and accuracy using the `sens.spec` command in mothur using the following equations.

@@

$$
F1_score = \frac{2 \times TP}{2 \times TP+FP+FN}
$$

$$
Accuracy = \frac{TP + TN}{(TP+FP+TN+FN)}
$$

To assess the reproducibility of the algorithms we randomized the starting order of each sequence collection ten times and ran each algorithm on each randomized collection. We then measured the MCC, F1 score, and accuracy for each randomization and quantified their coefficient of variation (CV; the ratio of the standard deviation to the mean).

To assess how the the memory and time requirements scaled with the number of sequences included in each sequence collection, we randomly subsampled 20, 40, 60, or 80% of the unique sequences in each collection. We obtained 10 subsamples at each depth for each dataset and ran each collection (N= 50 = 5 sequencing depths x 10 replicates) through each of the algorithms. We used the timeout script to quantify the maximum RAM used and the amount of time required to process each sequence collection (https://github.com/pshved/timeout). We limited each algorithm to 48 GB of RAM, 50 hours, and unless otherwise specified, a single processor.

***Data and code availability.*** The workflow utilized commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur [v.1.39.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015]. A reproducible version of this manuscript and analysis is available at https://github.com/SchlossLab/Westcott_OptiClust_mSystems_2015.

>>>>

***Sequence data and processing steps.***
To evaluate the OptiClust and the other algorithms we created two synthetic sequence collections and four sequence collections generated from previously published studies. The V4 region of the 16S rRNA gene was used from all datasets because it is a popular region that can be fully sequenced with two-fold coverage using the commonly used MiSeq sequencer from Illumina [@Kozich2013]. The method for generating the simulated datasets followed the approach used by Kopylova et al. [-@Kopylova2016] and Schloss [@Schloss2016]. Briefly, we randomly selected 10,000 uniques V4 fragments from 16S rRNA gene sequences that were unique from the SILVA non-redundant database [@Pruesse2007]. A community with an even relative abundance profile was generated by specifying that each sequence had a frequency of 100 reads. A community with a staggered relative abundance profile was generated by specifying that the abundance of each sequence was a randomly drawn integer sampled from a uniform distribution between 1 and 200. Sequence collections collected from human feces [@Baxter2016], murine feces [@Schloss2012], soil [@Johnston2016], and seawater [@Henson2016] were used to characterize the algorithms' performance with natural communities. These sequence collections were all generated using paired 150 or 250 nt reads of the V4 region. We re-processed all of the reads using a common analysis pipeline that included quality score-based error correction [@Kozich2013], alignment against a SILVA reference database [@Pruesse2007; @Schloss2010], screening for chimeras using UCHIME [@Edgar2011], and classification using a naive Bayesian classifier with the RDP training set requiring an 80% confidence score [@Wang2007].

***Implementation of clustering algorithms.***
In addition to the OptiClust algorithm we evaluated ten different *de novo* clustering algorithms. These included three hierarchical algorithms, average neighbor (AN), nearest neighbor (NN), and furthest neighbor (FN), which are implemented in mothur [v.1.39.0; @Schloss2009]. Seven heuristic methods were also used including abundance-based greedy clustering (AGC) and (distance-based greedy clustering) DGC as implemented in USEARCH [v.6.1; @Edgar2010] and VSEARCH [v.X.X.X; @Rognes2015], OTUClust [v.X.X.X; @XXXXXX], SumaClust [v.X.X.X; @XXXXXX], and Swarm [v.2.1.1; @Mah2014]. With the exception of Swarm each of these methods uses distance-based thresholds to report OTU assignments.

***Benchmarking.*** We evaluated the quality of the sequence clustering, reproducibility of the clustering, the speed of clustering, and the amount of memory required to complete the clustering. To assess the quality of the clusters generated by each method, we counted the cells within a confusion matrix that indicated how well the clusterings represented the distances between the pair of sequences [@Schloss2011Assessing]. Pairs of sequences that were in the same OTU and had a distance less than 3% were true positives (TPs), those that were in different OTUs and had a distance greater than 3% were true negatives (TNs), those that were in the same OTU and had a distance greater than 3% were false positives (FPs), and those that were in different OTUs and had a distance less than 3% were false negatives (FNs). To synthesize the matrix into a single metric we used the Matthews Correlation Coefficient using the `sens.spec` command in mothur using the following equations.

@@

To assess the reproducibility of the algorithms we randomized the starting order of each sequence collection ten times and ran each algorithm on each randomized collection. We then measured the MCC for each randomization and quantified their percent coefficient of variation (% CV; 100 times the ratio of the standard deviation to the mean).

To assess how the the memory and time requirements scaled with the number of sequences included in each sequence collection, we randomly subsampled 20, 40, 60, or 80% of the unique sequences in each collection. We obtained 10 subsamples at each depth for each dataset and ran each collection (N= 50 = 5 sequencing depths x 10 replicates) through each of the algorithms. We used the timeout script to quantify the maximum RAM used and the amount of time required to process each sequence collection (https://github.com/pshved/timeout). We limited each algorithm to 48 GB of RAM and 50 hours using a single processor.

***Data and code availability.*** The workflow utilized commands in GNU make (v.3.81), GNU bash (v.4.1.2), mothur [v.1.39.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015]. Within R we utilized the wesanderson [v.X.X.X; @XXXXXX], dplyr [v.X.X.X; @XXXXXX], tidyr [v.X.X.X; @XXXXXX], cowplot [v.X.X.X; @XXXXXX], and ggplot2 [v.X.X.X; @XXXXXX] packages. A reproducible version of this manuscript and analysis is available at https://github.com/SchlossLab/Westcott_OptiClust_mSystems_2015.
